Inattentive responding can induce spurious associations between task behavior and symptom measures

### 2023-04-01 manuscripts/main.tex
checkpointing revisions progress

% define highlight style
% \colorlet{usercolorname}{yellow!00}    % use for public version
\colorlet{usercolorname}{yellow!30}    % use for reviewer version
\sethlcolor{usercolorname}

@@

% setup references
\defbibheading{main}{}
\defbibheading{supp}{}
\addbibresource[label=main]{main.bib}
\addbibresource[label=supp]{supp.bib}
\renewcommand*{\bibfont}{\small}

@@

% set up the author block
\setlength{\affilsep}{2em}   % set the space between author and affiliation
\author[1]{Samuel Zorowitz}
\author[2]{Johanne Solis}
\author[1,3]{Yael Niv}
\author[4]{Daniel Bennett}
\affil[1]{Princeton Neuroscience Institute, Princeton University, NJ, USA}
\affil[2]{Rutgers-Princeton Center for Computational Cognitive Neuropsychiatry, Rutgers University, NJ, USA}
\affil[3]{Department of Psychology, Princeton University, NJ, USA}
\affil[4]{School of Psychological Sciences, Monash University, Victoria, Australia}

@@

% make abstract
\abstract{
A common research design in the field of computational psychiatry involves leveraging the power of online participant recruitment to assess correlations between behavior in cognitive tasks and the self-reported severity of psychiatric symptoms in large, diverse samples. Although large online samples have many advantages for psychiatric research, some potential pitfalls of this research design are not widely understood. Here we detail circumstances in which entirely spurious correlations may arise between task behavior and symptom severity as a result of inadequate screening of careless or low-effort responding on psychiatric symptom surveys. Specifically, since many psychiatric symptom surveys have asymmetric ground-truth score distributions in the general population, participants who respond carelessly on these surveys will show apparently elevated symptom levels. If these participants are similarly careless in their task performance, and are not excluded from analysis, this may result in a spurious association between greater symptom scores and worse behavioral task performance. Here, we demonstrate exactly this pattern of results in \hl{two independent samples of participants (total $N = 779$)} recruited online to complete a self-report symptom battery and \hl{one of two common cognitive tasks}. We show that many behavior-symptom correlations are entirely abolished when participants flagged for careless responding on surveys are excluded from analysis. We also show that exclusion based on task performance alone is \hl{often insufficient} to prevent these spurious correlations. Of note, we demonstrate that false-positive rates for these spurious correlations \emph{increase} with sample size, contrary to common assumptions. We offer guidance on how researchers using this general experimental design can guard against this issue in future research; in particular, we recommend the adoption of screening methods for self-report measures that are currently uncommon in this field.
}

@@

\begin{refsection}[main]
\section{Introduction}

In recent years, online labour markets (e.g. Amazon Mechanical Turk, Prolific, CloudResearch) have become increasingly popular as a source of research participants in the behavioral sciences \cite{stewart2017crowdsourcing}, in no small part due to the ease with which these services allow for recruitment of large, diverse samples. The advantages of online data collection have also begun to be recognized in psychiatric research \cite{chandler2016conducting}, where this method offers several distinct advantages over traditional approaches to participant recruitment. The ability to assess psychiatric symptom severity in large general-population samples makes possible large-scale transdiagnostic analysis \cite{gillan2016taking, rutledge2019machine}, and facilitates recruitment from difficult-to-reach participant populations \cite{strickland2019use}. Online labour markets also facilitate re-recruitment, making them an attractive option for validating the psychometric properties of assessment tools \cite{enkavi2019large} or studying clinical processes longitudinally \cite{kothe2019retention}.

With the advantages of online data collection also come specific drawbacks. Since participants recruited from online labour markets are typically completing experiments in their homes, they may be more likely to be distracted or multi-tasking during an experiment. They may also be more likely to use heuristic response strategies with the intention to minimise expenditure of time and cognitive effort (e.g., responding randomly on self-report surveys or behavioral tasks). Here, we will refer to such inattentive or low-effort behaviors as careless/insufficient effort (C/IE) responding \cite{huang2012detecting, curran2016methods}. Among researchers using online labour markets, a common view is that poor-quality data resulting from C/IE responding can simply be treated as a source of unsystematic measurement error that can be overcome with increased sample sizes \cite{gillan2016taking, chandler2020participant}. Common practice in online behavioral research is to mitigate poor-quality data using the same screening methods that are typically used in in-person data collection (e.g., excluding participants who perform at- or below-chance on behavioral tasks). However, these methods may be specifically inappropriate for online psychiatry studies, as we detail below.

Here we wish to draw special attention to an underappreciated feature of psychiatric research using self-report symptom surveys. In such surveys, participants rate their endorsement of various psychiatric symptoms and, since most individuals in the general population tend to endorse no or few symptoms in many symptom domains, the resulting ground-truth symptom distributions tend to be heavily positively skewed \cite{lowe2008validation, tomitaka2018distributional}. In this situation, the assumption that C/IE responding merely increases unsystematic measurement noise becomes untenable. Because of the positive skew in the ground-truth symptom distribution, participants who respond carelessly to the symptom survey are more likely to report higher levels of symptom endorsement relative to participants who complete the survey \hl{attentively} \cite{chandler2020participant, ophir2020turker, king2018random}. Consequently, unless C/IE survey responses are carefully identified and removed, a considerable proportion of putatively symptomatic individuals in an online sample may, in fact, be participants who have not engaged with the experiment with sufficient attention or effort.

@@

\begin{figure}[t]
\includegraphics[width=16cm]{../figures/main_01.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Simulated example of how spurious behavior-symptom correlations can arise when symptom endorsement is rare. \textit{Left:} When symptoms are moderately common in the general population, C/IE respondents (blue) are indistinguishable from attentive participants (red) in self-report measures (x-axis, marginal distribution shown on top). Despite the lower task performance of C/IE respondents (y-axis), no correlation arises between symptom scores and task performance (dots are participants drawn from the shown distributions, with 15\% C/IE participants; dashed line shows the (lack of) correlation.) \textit{Right:} When symptoms are rare in the general population, careless respondents appear symptomatic in self-report measures. As a result, self-report symptom scores show a significant correlation with task performance.}
\label{fig:simulation}
\end{figure}

Here, we demonstrate the real risk that C/IE responding can lead to spurious symptom-task correlations in computational psychiatry research. First, we asked to what extent recent studies in computational psychiatry screen participants based on self-report symptom data. We found that the majority of these studies did not screen participants' survey data at all, and that very few followed best-practice recommendations for survey data screening. We then asked whether behavioral screening alone was sufficient to identify participants engaging in C/IE responding on psychiatric symptom surveys. \hl{In two new datasets} from two separate online labour marketplaces, we found that screening based on task behavior fails to completely identify participants engaging in C/IE responding on surveys. Lastly, we investigated whether, under these circumstances, C/IE responding led to spurious correlations between symptom severity and task performance for positively-skewed symptom measures. Consistent with the logic set out above, we confirmed that failure to appropriately screen out C/IE survey responding in the proof-of-concept datasets that we collected would have produced a number of spurious correlations between task behavior and self-reported symptoms that are abolished when data are screened more thoroughly.

\section{Results}

\subsection{Screening for C/IE responding is common for task behavior but not for self-report surveys}

@@

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{8pt}
\begin{tabular}{lllrllr}
\toprule
& \multicolumn{2}{c}{Task Screening} & \multicolumn{4}{c}{Self-Report Screening} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-7}
Frequency & \multicolumn{2}{c}{N=39 (80\%)} & \multicolumn{4}{c}{N=19 (39\%)} \\
\midrule
Measure & Accuracy & 18 (37\%) & \multicolumn{2}{l}{Attention Check} & & 17 (35\%) \\
& Variability & 15 (31\%) & & Instructed & & 10 (20\%) \\
& Response Time & \ 7 (14\%) & & Infrequency & & 2 \, (4\%) \\
& Comprehension Check & \ \ 5 (10\%) & & Unspecified & & 5 (10\%) \\
& Other & 16 (33\%) & \multicolumn{2}{l}{Unobtrusive} & & \ 4 \, (8\%) \\
\bottomrule
\end{tabular}
\captionsetup{width=0.88\textwidth}
\caption{The prevalence and types of task and self-report data screening practices in a sample (N=49) of recent online behavioral studies.}
\label{tab:screening}
\end{table}

\subsection{Careless participants appear symptomatic when the overall level of symptom endorsement is low}

To measure the correspondence of screening measures estimated from task and self-report behavior, we conducted an online behavioral experiment involving a simple decision-making task and a battery of commonly used self-report psychiatric symptom measures (see Methods). A final sample of 386 participants from the Amazon Mechanical Turk (N=186) and Prolific (N=200) online labour markets completed a probabilistic reversal-learning task and 5 self-report symptom measures. The reversal-learning task required participants to learn through trial-and-error which of three options yielded reward most often, and was modeled after similar tasks used to probe reinforcement-learning deficits in psychiatric disorders \cite{waltz2007probabilistic, mukherjee_reward_2020}. The five self-report measures were the 7-up (which measures symptoms of hypomania), the 7-down (which measures symptoms of depression), the GAD-7, (which measures generalized anxiety symptoms), the BIS/BAS (which measures reward and punishment motivations), the SHAPS (which measures anhedonia symptoms), and the PSWQ (which measures worry symptoms), and were chosen based on previous literature to have a variety of expected response distributions (symmetric and asymmetric). In line with current best-practice recommendations in personality psychology \cite{huang2015detecting}, each self-report instrument included one `infrequency' item that could be used to identify C/IE responses in survey data (see Methods for a list of infrequency items). The entire experiment (surveys and task) was designed to require 10 minutes on average to complete (observed mean = 10.28 minutes). To minimize any influence of fatigue on survey responding, participants completed the surveys prior to beginning the task.

To assess the overall quality of the data, we examined the number of participants flagged by the choice accuracy and infrequency item screening measures. Only 26 participants (7\%) were flagged as exhibiting choice behavior at or below statistically chance levels in the reversal-learning task. In contrast, 85 participants (22\%) endorsed a logically invalid or improbable response on one or more of the infrequency items when completing the self-report symptom measures. This discrepancy in the proportion of participants flagged by each method is consistent with previous research, which found varying levels of sensitivity to C/IE responding across screening methods \cite{desimone2018dirty}. The proportion of participants flagged for C/IE responding was marginally but significantly greater on Mechanical Turk compared to Prolific for both task (MTurk: N=18/186; Prolific: N=8/200; $z=2.22, p=0.026$) and survey data (MTurk: 50/186; Prolific: 35/200; $z = 2.22, p = 0.026$).

\begin{figure}[t!]
\includegraphics[width=16cm]{../figures/main_02a.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Raincloud plots of total symptom scores in attentive (red) and C/IE (blue) participants. Each colored dot represents the symptom score for one participant. Black circles: average score within each group (error bars denote 95\% bootstrap confidence interval). Shaded plots: estimated distribution of responses for each group of participants. The scales are ordered approximately according to their estimated skew (see Table 2) from top-left (7-up) to bottom-right (PSWQ). The average level of symptom endorsement is most markedly different between groups in measures with lowest overall rates of endorsement.}
\label{fig:distributions}
\end{figure}

We hypothesise that spurious behavior-symptom correlations may emerge due to a mean-shift in the average level of symptom endorsement in participants engaging in C/IE responding relative to attentive participants. In turn, a mean-shift is expected to occur when the overall rate of symptom endorsement is low; \hl{that is, comparably higher scores are more likely for C/IE participants responding at random on a right skewed questionnaire.} In line with our predictions, the average level of symptom endorsement was noticeably exaggerated in C/IE-responding participants for the symptom measures where symptom scores were most positively-skewed (7-up, 7-down, GAD-7; see Figure \ref{fig:distributions}). In contrast, where there was higher rates of symptom endorsement overall, the distributions of symptom scores between the two groups of participants were less noticeably distinct. Permutation testing confirmed that observed mean-shifts in symptom scores for C/IE participants were statistically significant for the majority of symptom measures (Table \ref{tab:validation}).

\begin{table}[!t]
\centering
\small
\setlength{\tabcolsep}{6pt}
\begin{tabular}{lcccccccc}
\toprule
& & \multicolumn{3}{c}{Total Score} & \multicolumn{2}{c}{Cronbach's $\alpha$} & \multicolumn{2}{c}{\% Clinical Cutoff} \\
\cmidrule(lr){3-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
 Subscale & Skew & Attentive & C/IE & \textit{t}-value & Attentive & C/IE & Before & After \\
\midrule
7-up   &  0.81 &    3.87 &   10.15 &  $-13.31^*$ &    0.84 &    0.84 &     13.0\% &       4.0\% \\
7-down &  0.76 &    4.75 &   10.68 &   $-9.99^*$ &    0.94 &    0.88 &     17.4\% &       9.3\% \\
GAD-7  &  0.75 &    4.86 &    9.73 &   $-7.88^*$ &    0.92 &    0.87 &     25.9\% &      17.3\% \\
BIS    &  0.78 &    7.74 &    7.92 &   -0.54 &    0.83 &    0.62 &        - &             - \\
BAS    &  0.17 &   15.73 &   16.16 &   -0.91 &    0.84 &    0.71 &        - &             - \\
SHAPS  &  0.26 &    8.02 &   10.85 &   $-4.04^*$ &    0.90 &    0.81 &     17.9\% &      14.6\% \\
PSWQ   &  0.19 &    4.78 &    6.74 &   $-4.78^*$ &    0.93 &    0.81 &      7.3\% &       7.0\% \\
\bottomrule
\end{tabular}
\captionsetup{width=0.88\textwidth}
\caption{Descriptive statistics of the self-report symptom measures between attentive and C/IE participants. Skew: the empirical skewness of the distribution of total symptom scores. Total score: the average symptom score across attentive and C/IE participants. Stars indicate statistical significance at $p<0.05$. Cronbach's $\alpha$: a measure of response consistency, where values closer to 1 indicate greater consistency in responses. \% Clinical Cutoff: the percentage of participants reaching threshold for clinical symptomology before and after screening based on the infrequency measure. (The BIS/BAS scales do not have clinical thresholds.)}
\label{tab:validation}
\end{table}

Hereafter, we use the infrequency-item method as a primary means of identifying C/IE responding in our data. To verify this approach, we conducted three validation analyses. The first analysis compared estimated internal consistency of self-report measures between the C/IE and attentive groups. The logic is that, if C/IE responding manifests as a tendency to respond randomly, we should expect to see a decrease in the consistency of a measure in the C/IE responding group \cite{maniaci2014caring, desimone2018dirty, desimone2018differential}. In line with this reasoning, we observed a reduction in Cronbach's $\alpha$ in the C/IE group for the majority of survey instruments (Table \ref{tab:validation}). A permutation test confirmed that the average decrease in internal consistency across measures was greater than would be expected by chance given the difference in participant numbers between groups ($t = 3.69, p = 0.021$).

Second, we quantified the degree to which participants responded to self-report symptom surveys in a stereotyped fashion; that is, we determined if participants exhibited patterns in their responses that were independent of the contents of the survey items. We fit a random-intercept item factor-analysis model \cite{maydeu2006random} to self-report data (see Methods), and for each participant we estimated an intercept parameter that quantified their bias towards using responses on the left or right side of the response scale, regardless of what that response signifies for a particular self-report measure (e.g., low on one symptom scale versus high on another). We observed a credible difference between the average value of this intercept for the two groups ($\Delta \text{intercept} = 0.67, 95\% \ \text{HDI} = [-0.78, -0.55]$), such that C/IE participants were biased towards using the right-half of survey response options. This translates to a tendency to endorse \emph{more severe} symptoms on the 7-up/7-down and GAD-7 scales (where the rightmost options indicate greater frequency of symptoms) but \emph{less extreme} symptoms or \hl{personality traits on the SHAPS and BIS (where the rightmost options indicate lower frequency of symptoms or personality traits)} despite these inventories measuring strongly correlated constructs (i.e. depression and anhedonia, anxiety and behavioral inhibition).

@@

\subsection{Low correspondence between task and self-report measures of C/IE responding}

@@

\begin{figure}[!t]
\includegraphics[width=16cm, trim=0 0 3cm 0]{../figures/main_02b.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Similarity of task and self-report data screening measures. Each tile corresponds to the Spearman rank correlation (left) and Dice similarity coefficient (right) between two screening measures. Acronyms: INF = infrequency item; ISD = inter-item standard deviation; REL = personal reliability; MAH = Mahalanobis distance; READ = reading time; VAR = choice variability; ACC = choice accuracy; WSLS = win-stay lose-shift rate; RT = suspicious response times. Similarity scores have been thresholded after correcting for multiple comparisons. Numbers denote the strength of statistically significant correlations. Cross-modality correlations between task-behavior (left) and infrequency-item self-report measures (bottom) are in the dashed rectangle.}
\label{fig:correspondence}
\end{figure}

@@

\subsection{Spurious symptom-behavior correlations produced by C/IE responding}

@@

For each participant, we computed both descriptive and computational-model-based measures of behavior on the reversal learning task (see Methods). To understand the effects of applying different forms of screening, we estimated the correlations between each unique pairing of a self-report symptom measure and measure of behavior under four different conditions: no screening, screening only on task behavior (i.e. only participants whose choice accuracy was above chance), screening only on self-report responses (i.e. only participants who responded correctly on all infrequency items), or both. The resulting pairwise behavior-symptom correlations following each screening procedure are presented in Figure \ref{fig:correlations}. We note that we did not correct these correlation analyses for multiple comparisons, since our purpose was to demonstrate the extent of this issue across multiple behavioral measures and self-report symptoms. Any one of these correlations considered individually can be thought of as emulating a conventional analysis where fewer statistical tests would be performed.

\begin{figure}[!t]
\includegraphics[width=17cm]{../figures/main_03a.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Absolute Spearman rank correlations between task behavior (y-axis) and symptom measures (x-axis) under different regimes of data screening and participant exclusions. Only statistically significant correlations are shown ($p<0.05$ not corrected for multiple comparisons; \hl{all signed correlations are shown in Supplementary Figure} \ref{fig:correlations_s1}). Black Xs indicate significant correlations ablated under screening. No Screening = no exclusions; Accuracy Only = exclusions based on chance-level performance in the reversal-learning task; Infrequency Only = exclusions based on invalid or improbable responses to infrequency items; Both Types = exclusions based on the previous two measures. Acronyms: Acc = choice accuracy; Pts = total points earned; WS = win-stay rate; LS = lose-shift rate; Pers = perseveration errors; $\beta$ = inverse temperature; $\eta_p$ = positive learning rate; $\eta_n$ = negative learning rate; $\kappa$ = learning rate asymmetry.}
\label{fig:correlations}
\end{figure}

When no rejections based on C/IE responding was applied (i.e. all participants were included in the analysis; Figure \ref{fig:correlations}A), many significant correlations emerged between measures of task behavior and symptom scores, in particular for 4 of the self-report instruments (7-up, which measures symptoms of hypomania; 7-down, which measures symptoms of depression; GAD-7, which measures generalized anxiety symptoms; and BIS, which measures tendencies related to behavioral inhibition). Consistent with our predictions, the majority of these correlations involved symptom measures with asymmetric total score distributions. Attending to only the most skewed measures (i.e. 7-up, 7-down, GAD-7), symptom endorsement was correlated with almost every behavioral measure. That is, significant correlations were not restricted only to general behavioral measures often used as proxies for participant effort (e.g. accuracy, inverse temperature $\beta$) but also to measures of specific theoretical interest, such as asymmetry of learning from positive and negative reward prediction errors ($\kappa$). Conversely, we found few significant correlations among symptom measures with more symmetric distributions. This is despite the fact these scales measure similar symptoms and syndromes (e.g. anxiety as measured by the GAD-7 and worry as measured by the PSWQ; depression as measured by the 7-down and anhedonia as measured by the SHAPS).

Next, we excluded participants from analysis based on task-behavior screening (i.e. choice accuracy, removing the 7\% of participants exhibiting behavior indistinguishable from chance; Figure \ref{fig:correlations}B). The pattern of correlations was largely unchanged: we again found many significant correlations between measures of behavior and asymmetric symptom measures, but almost no significant correlations involving symmetric symptom measures. This suggests that rejection of participants based on the most common form of behavioral screening (i.e. performance accuracy) had little effect on behavior-symptom correlations as compared to no screening.

In stark contrast, when we rejected participants based on self-report screening  (removing 22\% of participants who endorsed one or more invalid or improbable responses on the infrequency items; Figure \ref{fig:correlations}C), the number of significant correlations was markedly reduced, particularly for several of the most skewed symptom measures (7-down, GAD-7) and proxy measures of task attentiveness (e.g. accuracy, inverse temperature). This pattern of correlations was largely similar when rejections were applied based on both task and self-report screening measures (Figure \ref{fig:correlations}D). We also note that with stricter screening, the remaining significant correlations were, for the most part (but not always), weaker.

@@

\subsection{\texorpdfstring{\hl{Individuals with and without diagnosed psychiatric disorders fail attention checks at similar rates}}{}}

@@

\section{Discussion}

>>>>

@@

% setup references
\defbibheading{main}{}
\addbibresource[label=main]{main.bib}
\renewcommand*{\bibfont}{\small}

@@

% set up the author block
\setlength{\affilsep}{2em}   % set the space between author and affiliation
\author[1,*]{Samuel Zorowitz}
\author[2]{Johanne Solis}
\author[1,3]{Yael Niv}
\author[4]{Daniel Bennett}
\affil[1]{Princeton Neuroscience Institute, Princeton University, NJ, USA}
\affil[2]{Rutgers-Princeton Center for Computational Cognitive Neuropsychiatry, Rutgers University, NJ, USA}
\affil[3]{Department of Psychology, Princeton University, NJ, USA}
\affil[4]{School of Psychological Sciences, Monash University, Victoria, Australia}
\affil[*]{Corresponding author (zorowitz@princeton.edu)}

@@

% make abstract
\abstract{
A common research design in the field of computational psychiatry involves leveraging the power of online participant recruitment to assess correlations between behavior in cognitive tasks and the self-reported severity of psychiatric symptoms in large, diverse samples. Although large online samples have many advantages for psychiatric research, some potential pitfalls of this research design are not widely understood. Here we detail circumstances in which entirely spurious correlations may arise between task behavior and symptom severity as a result of inadequate screening of careless or low-effort responding on psychiatric symptom surveys. Specifically, since many psychiatric symptom surveys have asymmetric ground-truth score distributions in the general population, participants who respond carelessly on these surveys will show apparently elevated symptom levels. If these participants are similarly careless in their task performance, and are not excluded from analysis, this may result in a spurious association between greater symptom scores and worse behavioral task performance. Here, we demonstrate exactly this pattern of results in two independent samples of participants (total $N = 779$) recruited online to complete a self-report symptom battery and one of two common cognitive tasks. We show that many behavior-symptom correlations are entirely abolished when participants flagged for careless responding on surveys are excluded from analysis. We also show that exclusion based on task performance alone is often insufficient to prevent these spurious correlations. Of note, we demonstrate that false-positive rates for these spurious correlations \emph{increase} with sample size, contrary to common assumptions. We offer guidance on how researchers using this general experimental design can guard against this issue in future research; in particular, we recommend the adoption of screening methods for self-report measures that are currently uncommon in this field.
}

@@

\begin{refsection}[main]
\section*{Introduction}

In recent years, online labor markets (e.g., Amazon Mechanical Turk, Prolific, CloudResearch) have become increasingly popular as a source of research participants in the behavioral sciences \cite{stewart2017crowdsourcing}, in no small part due to the ease with which these services allow for recruitment of large, diverse samples. The advantages of online data collection have also begun to be recognized in psychiatric research \cite{chandler2016conducting}, where this method offers several distinct advantages over traditional approaches to participant recruitment. The ability to assess psychiatric symptom severity in large general-population samples makes possible large-scale transdiagnostic analysis \cite{gillan2016taking, rutledge2019machine}, and facilitates recruitment from difficult-to-reach participant populations \cite{strickland2019use}. Online labor markets also facilitate re-recruitment, making them an attractive option for validating the psychometric properties of assessment tools \cite{enkavi2019large} or studying clinical processes longitudinally \cite{kothe2019retention}.

With the advantages of online data collection also come specific drawbacks. Since participants recruited from online labor markets are typically completing experiments in their homes, they may be more likely to be distracted or multi-tasking during an experiment. They may also be more likely to use heuristic response strategies with the intention to minimise expenditure of time and cognitive effort (e.g., responding randomly on self-report surveys or behavioral tasks). Here, we will refer to such inattentive or low-effort behaviors as careless/insufficient effort (C/IE) responding \cite{huang2012detecting, curran2016methods}. Among researchers using online labor markets, a common view is that poor-quality data resulting from C/IE responding can simply be treated as a source of unsystematic measurement error that can be overcome with increased sample sizes \cite{gillan2016taking, chandler2020participant}. Common practice in online behavioral research is to mitigate poor-quality data using the same screening methods that are typically used in in-person data collection (e.g., excluding participants who perform at- or below-chance on behavioral tasks). However, these methods may be specifically inappropriate for online psychiatry studies, as we detail below.

Here we wish to draw special attention to an underappreciated feature of psychiatric research using self-report symptom surveys. In such surveys, participants rate their endorsement of various psychiatric symptoms and, since most individuals in the general population tend to endorse no or few symptoms in many symptom domains, the resulting ground-truth symptom distributions tend to be heavily positively skewed \cite{lowe2008validation, tomitaka2018distributional}. In this situation, the assumption that C/IE responding merely increases unsystematic measurement noise becomes untenable. Because of the positive skew in the ground-truth symptom distribution, participants who respond carelessly to the symptom survey are more likely to report higher levels of symptom endorsement relative to participants who complete the survey attentively \cite{chandler2020participant, ophir2020turker, king2018random}. Consequently, unless C/IE survey responses are carefully identified and removed, a considerable proportion of putatively symptomatic individuals in an online sample may, in fact, be participants who have not engaged with the experiment with sufficient attention or effort.

@@

Here, we demonstrate the real risk that C/IE responding can lead to spurious symptom-task correlations in computational psychiatry research. First, we asked to what extent recent studies in computational psychiatry screen participants based on self-report symptom data. We found that the majority of these studies did not screen participants' survey data at all, and that very few followed best-practice recommendations for survey data screening. We then asked whether behavioral screening alone was sufficient to identify participants engaging in C/IE responding on psychiatric symptom surveys. In two new datasets from two separate online labor marketplaces, we found that screening based on task behavior fails to completely identify participants engaging in C/IE responding on surveys. Lastly, we investigated whether, under these circumstances, C/IE responding led to spurious correlations between symptom severity and task performance for positively-skewed symptom measures. Consistent with the logic set out above, we confirmed that failure to appropriately screen out C/IE survey responding in the proof-of-concept datasets that we collected would have produced a number of spurious correlations between task behavior and self-reported symptoms that are abolished when data are screened more thoroughly.

\section*{Results}

\subsection*{Screening for C/IE responding is common for task behavior but not for self-report surveys}

@@

\subsection*{Careless participants appear symptomatic when the overall level of symptom endorsement is low}

To measure the correspondence of screening measures estimated from task and self-report behavior, we conducted an online behavioral experiment involving a simple decision-making task and a battery of commonly used self-report psychiatric symptom measures (see Methods). A final sample of 386 participants from the Amazon Mechanical Turk (N=186) and Prolific (N=200) online labor markets completed a probabilistic reversal-learning task and 5 self-report symptom measures. The reversal-learning task required participants to learn through trial-and-error which of three options yielded reward most often, and was modeled after similar tasks used to probe reinforcement-learning deficits in psychiatric disorders \cite{waltz2007probabilistic, mukherjee_reward_2020}. The five self-report measures were the 7-up (which measures symptoms of hypomania), the 7-down (which measures symptoms of depression), the GAD-7, (which measures generalized anxiety symptoms), the BIS/BAS (which measures reward and punishment motivations), the SHAPS (which measures anhedonia symptoms), and the PSWQ (which measures worry symptoms), and were chosen based on previous literature to have a variety of expected response distributions (symmetric and asymmetric). In line with current best-practice recommendations in personality psychology \cite{huang2015detecting}, each self-report instrument included one `infrequency' item that could be used to identify C/IE responses in survey data (see Methods for a list of infrequency items). The entire experiment (surveys and task) was designed to require 10 minutes on average to complete (observed mean = 10.28 minutes). To minimize any influence of fatigue on survey responding, participants completed the surveys prior to beginning the task.

To assess the overall quality of the data, we examined the number of participants flagged by the choice accuracy and infrequency item screening measures. Only 26 participants (7\%) were flagged as exhibiting choice behavior at or below statistically chance levels in the reversal-learning task. In contrast, 85 participants (22\%) endorsed a logically invalid or improbable response on one or more of the infrequency items when completing the self-report symptom measures. This discrepancy in the proportion of participants flagged by each method is consistent with previous research, which found varying levels of sensitivity to C/IE responding across screening methods \cite{desimone2018dirty}. The proportion of participants flagged for C/IE responding was marginally but significantly greater on Mechanical Turk compared to Prolific for both task (MTurk: N=18/186; Prolific: N=8/200; two-tailed, two-sample proportions test: $z(384)=2.224$, $p=0.026$, $h=0.230$, 95\% CI = [0.006, 0.107]) and survey data (MTurk: 50/186; Prolific: 35/200; two-tailed, two-sample proportions test: $z(384)=2.223$, $p=0.026$, $h=0.227$, 95\% CI = [0.011, 0.176]).

We hypothesise that spurious behavior-symptom correlations may emerge due to a mean-shift in the average level of symptom endorsement in participants engaging in C/IE responding relative to attentive participants. In turn, a mean-shift is expected to occur when the overall rate of symptom endorsement is low; that is, comparably higher scores are more likely for C/IE participants responding at random on a right skewed questionnaire. In line with our predictions, the average level of symptom endorsement was noticeably exaggerated in C/IE-responding participants for the symptom measures where symptom scores were most positively-skewed (7-up, 7-down, GAD-7; see Figure \ref{fig:distributions}). In contrast, where there was higher rates of symptom endorsement overall, the distributions of symptom scores between the two groups of participants were less noticeably distinct. Permutation testing confirmed that observed mean-shifts in symptom scores for C/IE participants were statistically significant for the majority of symptom measures (Table \ref{tab:validation}).

Hereafter, we use the infrequency-item method as a primary means of identifying C/IE responding in our data. To verify this approach, we conducted three validation analyses. The first analysis compared estimated internal consistency of self-report measures between the C/IE and attentive groups. The logic is that, if C/IE responding manifests as a tendency to respond randomly, we should expect to see a decrease in the consistency of a measure in the C/IE responding group \cite{maniaci2014caring, desimone2018dirty, desimone2018differential}. In line with this reasoning, we observed a reduction in Cronbach's $\alpha$ in the C/IE group for the majority of survey instruments (Table \ref{tab:validation}). A permutation test confirmed that the average decrease in internal consistency across measures was greater than would be expected by chance given the difference in participant numbers between groups (two-tailed, paired-samples $t$-test: $t(6)=3.689$, $p=0.021$, $d=1.506$, 95\% CI = [0.048, 0.141]).

Second, we quantified the degree to which participants responded to self-report symptom surveys in a stereotyped fashion; that is, we determined if participants exhibited patterns in their responses that were independent of the contents of the survey items. We fit a random-intercept item factor-analysis model \cite{maydeu2006random} to self-report data (see Methods), and for each participant we estimated an intercept parameter that quantified their bias towards using responses on the left or right side of the response scale, regardless of what that response signifies for a particular self-report measure (e.g., low on one symptom scale versus high on another). We observed a credible difference between the average value of this intercept for the two groups ($\Delta \text{intercept} = 0.67, 95\% \ \text{HDI} = [-0.78, -0.55]$), such that C/IE participants were biased towards using the right-half of survey response options. This translates to a tendency to endorse \emph{more severe} symptoms on the 7-up/7-down and GAD-7 scales (where the rightmost options indicate greater frequency of symptoms) but \emph{less extreme} symptoms or personality traits on the SHAPS and BIS (where the rightmost options indicate lower frequency of symptoms or personality traits) despite these inventories measuring strongly correlated constructs (i.e., depression and anhedonia, anxiety and behavioral inhibition).

@@

\subsection*{Low correspondence between task and self-report measures of C/IE responding}

@@

@@

\subsection*{Spurious symptom-behavior correlations produced by C/IE responding}

@@

For each participant, we computed both descriptive and computational-model-based measures of behavior on the reversal learning task (see Methods). To understand the effects of applying different forms of screening, we estimated the correlations between each unique pairing of a self-report symptom measure and measure of behavior under four different conditions: no screening, screening only on task behavior (i.e., only participants whose choice accuracy was above chance), screening only on self-report responses (i.e., only participants who responded correctly on all infrequency items), or both. The resulting pairwise behavior-symptom correlations following each screening procedure are presented in Figure \ref{fig:correlations}. We note that we did not correct these correlation analyses for multiple comparisons, since our purpose was to demonstrate the extent of this issue across multiple behavioral measures and self-report symptoms. Any one of these correlations considered individually can be thought of as emulating a conventional analysis where fewer statistical tests would be performed.

When no rejections based on C/IE responding was applied (i.e., all participants were included in the analysis; Figure \ref{fig:correlations}A), many significant correlations emerged between measures of task behavior and symptom scores, in particular for 4 of the self-report instruments (7-up, which measures symptoms of hypomania; 7-down, which measures symptoms of depression; GAD-7, which measures generalized anxiety symptoms; and BIS, which measures tendencies related to behavioral inhibition). Consistent with our predictions, the majority of these correlations involved symptom measures with asymmetric total score distributions. Attending to only the most skewed measures (i.e., 7-up, 7-down, GAD-7), symptom endorsement was correlated with almost every behavioral measure. That is, significant correlations were not restricted only to general behavioral measures often used as proxies for participant effort (e.g., accuracy, inverse temperature $\beta$) but also to measures of specific theoretical interest, such as asymmetry of learning from positive and negative reward prediction errors ($\kappa$). Conversely, we found few significant correlations among symptom measures with more symmetric distributions. This is despite the fact these scales measure similar symptoms and syndromes (e.g., anxiety as measured by the GAD-7 and worry as measured by the PSWQ; depression as measured by the 7-down and anhedonia as measured by the SHAPS).

Next, we excluded participants from analysis based on task-behavior screening (i.e., choice accuracy, removing the 7\% of participants exhibiting behavior indistinguishable from chance; Figure \ref{fig:correlations}B). The pattern of correlations was largely unchanged: we again found many significant correlations between measures of behavior and asymmetric symptom measures, but almost no significant correlations involving symmetric symptom measures. This suggests that rejection of participants based on the most common form of behavioral screening (i.e., performance accuracy) had little effect on behavior-symptom correlations as compared to no screening.

In stark contrast, when we rejected participants based on self-report screening  (removing 22\% of participants who endorsed one or more invalid or improbable responses on the infrequency items; Figure \ref{fig:correlations}C), the number of significant correlations was markedly reduced, particularly for several of the most skewed symptom measures (7-down, GAD-7) and proxy measures of task attentiveness (e.g., accuracy, inverse temperature). This pattern of correlations was largely similar when rejections were applied based on both task and self-report screening measures (Figure \ref{fig:correlations}D). We also note that with stricter screening, the remaining significant correlations were, for the most part (but not always), weaker.

@@

\subsection*{Individuals with and without diagnosed psychiatric disorders fail attention checks at similar rates}

@@

\section*{Discussion}
